Modularly Composing Typed Language Fragments

### 2014-11-04 att-pldi15/att-pldi15.tex
updated

In this paper, we begin from first principles by specifying a language, @$\lambda$, in the style of many full-scale languages: as a bidirectionally typed translation semantics. Only the $\rightarrow$ type constructor (tycon) is built in; all other external tycons (we show constrained strings and a variant on records) are defined by extending a \emph{tycon context}. Each tycon defines the semantics of its  term-level operators (e.g. record projection) using a static language where types and translations are values. The semantics includes a simple check that leads to powerful metatheoretic guarantees, notably \emph{type safety} and \emph{conservativity}: that  \emph{tycon invariants} will always be conserved under extension. Type system providers need not provide mechanized proofs. Instead, the check enforces a \emph{translation independence} property via type abstraction, the same principle underlying ML-style module systems.
\end{abstract}

@@

Universality properties (e.g. ``Turing-completeness'') are often invoked to guarantee that an embedding that preserves a desirable fragment's dynamics can be constructed, but an isomorphic embedding must also preserve the static semantics and, if defined, performance bounds specified using a cost semantics. This is not always possible. 
Embeddings are also sometimes too \emph{complex}, as measured by the cost of the extralinguistic computations that are needed to map in and out of the embedding and, if these  must be performed mentally by programmers, considering various human factors. %We will discuss specific examples below.
%When an embedding is too complex, abstraction providers have a few options, discussed in Sec. \ref{prior-work}. When an embedding is not possible, however, or when these options are  insufficient, 
%providers are often compelled to introduce these fragments by extending an existing language, thereby forming a new \emph{dialect}. To save effort, they may do so by forking existing artifacts or leverage tools like compiler generators or language frameworks, also reviewed in Sec. \ref{prior-work}. %Indeed, the proliferation of language dialects constructed for this reason might be taken as an evidence that the core language is not as ``general'', when considered comprehensively, as might be hoped. 
Each time a fragment must be exposed directly, rather than as a library, a new \emph{dialect} of the language must be constructed. Within the ML lineage, for example, dialects that go beyond ``core ML'' abound:
%Reynolds, in a remark that recalls the ``Turing tarpit'' of Perlis \cite{Perl82a}, summarizes the issue \cite{Reynolds94anintroduction}: 
%\begin{quote}
%To say that any reasonable function can be expressed by some program is not to say that it can be expressed by the most reasonable program. It is clear that the language requires a novel programming style. Moreover, it is likely that certain important functions cannot be expressed by their most efficient algorithms.
%\end{quote}
%

\begin{enumerate}
\vspace{-5px}
%\compresslist
\item 
\textbf{General Purpose Fragments:} 
A number of variations on product types, for example, have been introduced in dialects: 
$n$-ary tuples, 
labeled tuples, 
records (identified up to reordering), 
records with width and depth subtyping \cite{Cardelli:1984:SMI:1096.1098}, 
records with update operators\footnote{The Haskell wiki notes that ''No, extensible records are not implemented in GHC. The problem is that the record design space is large, and seems to lack local optima. [...] As a result, nothing much happens.'' \cite{GHCFAQ}} \cite{ocaml-manual}, 
records with mutable fields \cite{ocaml-manual}, 
%field delegation \cite{atlang-gpce14} \todo{gpce submission} 
and 
records with ``methods'' (i.e. pure objects \cite{TSLs}). 
 Sum-like types are also exposed in various ways: 
finite datatypes, 
open datatypes \cite{conf/ppdp/LohH06}, 
hierarchically open datatypes \cite{journals/toplas/MillsteinBC04}, 
polymorphic variants \cite{ocaml-manual} and 
ML-style exception types. Combinations of these manifest themselves as class-based object systems \cite{ocaml-manual}.

\item
\textbf{Specialized Fragments:} Fragments that track specialized static invariants to provide stronger correctness or security guarantees, manage unwieldy lower-level abstractions and run-time systems or control cost are also frequently introduced in dialects, e.g. for data parallelism  \cite{chakravarty2007data}, distributed programming \cite{Murphy:2007:TDP:1793574.1793585}, reactive programming \cite{mandel2005reactiveml}, authenticated data structures \cite{Miller:2014:ADS:2535838.2535851}, databases \cite{Ohori:2011:MSM:2034773.2034815},  units of measure \cite{conf/cefp/Kennedy09} and regular string sanitation \cite{sanitation-psp14}. %\verb|sprintf| in the OCaml dialect statically distinguishes format strings from strings.% All of these are implemented as dialects of existing languages, presumably because a strong encoding was not feasible.
\vspace{-3px}
\item
\textbf{Foreign Fragments:} A safe and natural foreign function interface (FFI) can be valuable (particularly given this proliferation of dialects). This requires enforcing the type system of the foreign language in the calling language. %Using  a FFI that does not do this can lead to safety issues, even when both languages are separately known to be safe. 
%Safe FFIs generally require direct extensions to the language. 
For example, MLj builds in a safe FFI to Java \cite{Benton:1999:IWW:317636.317791}.
\end{enumerate}
\vspace{-5px}

This \emph{dialect-oriented} state of affairs is unsatisfying. %Language designers are burdened with needing to understand the complete metatheory of every fragment they add to their language to make sure it is not incompatible with existing fragments, so languages change rarely. This decreases the impact of many potentially useful innovations designed by fragment providers (e.g. many academic researchers) by making them unavailable to programmers. 
While programmers can choose from dialects supporting, e.g., a principled approach to distributed programming, or one that builds in support for statically reasoning about units of measure, one that supports both fragments may not be available. Using different dialects separately for different components of a program is untenable: components written in different dialects cannot always interface safely (i.e. a safe FFI, item 3 above, is needed between every pair of dialects). And as just mentioned, composing dialects is non-trivial.

These problems do not arise for a fragment that can be expressed as an isomorphic embedding (i.e. as a library) because modern \emph{module systems} can enforce abstraction barriers that ensure that the isomorphism needs only to be established in the ``closed world'' of the module. For example, a module defining sets in ML can hold the representation of sets abstract, ensuring that any invariants maintained by the functions in the module (e.g. uniqueness, if using a list representation) will hold no matter which other modules are in use by a client \cite{harper1997programming}. Other languages provide similar facilities, e.g. Scala's abstract type members \cite{conf/oopsla/AminRO14}. %Mechanisms that can help decrease the complexity of an embedding without violating abstraction barriers are thus valuable, and we will lead into our work in Sec. \ref{prior-work} by summarizing them.

When library-based embeddings are not possible, as in the examples above, mechanisms are needed that make it possible to define and  reason in a similarly modular manner about  direct extensions to the semantics of a language. Such a mechanism could ultimately be integrated directly into the language, blurring the distinction between fragments and libraries and decreasing the need for new dialects.% Importing fragments that introduce new semantics   would be as easy and reasonable as importing a new module is in ML today. %In the limit, the community could rely on modularly mechanized metatheory and compiler correctness results.% rather than requiring heroic efforts from individual research groups that consider an entire language at once. %Recent work has shown progress on modularly introducing new concrete syntax, reviewed in Sec. \ref{desugaring}. Our focus is on the problem of introducing new semantics.

\paragraph{Contributions} In this paper, we take foundational steps toward this goal by  constructing a simple but surprisingly powerful core calculus, @$\lambda$ (the ``actively typed'' lambda calculus). %Despite its minimality, it can host a variety of practical semantic extensions like those described above, while maintaining strong metatheoretic guarantees and, crucially, providing modular reasoning properties. %a calculus introduced briefly in recent work on \emph{active type constructors}  \cite{atlang-gpce14}\todo{TR? Arxiv?}, reviewed in Sec. \ref{overview}. 
Its semantics are structured like those of many modern languages, consisting of an \emph{external language} (EL) governed by a {typed translation semantics} targeting a much simpler \emph{internal language} (IL). 
Rather than building in a monolithic set of external type constructors, however, the  semantics are indexed by a \emph{tycon context}. Each tycon defines the semantics of its operators via functions written in a \emph{static language} (SL) where types and translations are values. %, which provides a form of type-level computation. %The authors demonstrate the expressive power of this technique primarily with an implementation and a number of examples of fragments as libraries, and their core calculus .% (unlike systems that treat the type system as a ``bag of rules'', where non-determinism can arise).

We will begin by giving an overview of the organization  and main judgements of @$\lambda$ in Sec. \ref{atlam}, then discuss how types are constructed and introduce our two main examples, one defining labeled product types with a functional update operator, and the other regular string types, based on a recent core calculus style specification \cite{sanitation-psp14}, in Sec. \ref{types}. We describe how tycons control the semantics of their term-level operator constructors (opcons) in Sec. \ref{external-terms}.  We next give the key metatheoretic properties of the calculus in Sec. \ref{metatheory}, including \emph{type safety} and a key modularity result, which we call  \emph{conservativity}: any invariants that can be established about all values of a type under \emph{some} tycon context (i.e. in some  ``closed world'') are conserved in any further extended tycon context (i.e. in the ``open world''). Interestingly, type system providers need not provide mechanized proofs to maintain these guarantees. Instead, the approach we take relies on type abstraction in the internal language. As a result, we are able to use the same parametricity results that underly modular reasoning in simply-typed languages like ML to reason modularly about typed language fragments. We conclude with  related and future work in Sec. \ref{prior-work}.

\section{Overview of @$\lambda$}\label{atlam}\label{overview}
\begin{figure}[t]
\small
\hspace{-5px}\input{syntax-table-IL}
\caption{Syntax of {$\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\iunit}\,{\times}\,{+}\}$}, our internal language (IL). Metavariable $x$ ranges over term variables and $\alpha$ and $t$ both range over type variables.}
\label{syntax-IL}
\vspace{-10px}
\end{figure}
\paragraph{External Language}
Programmers interface with @$\lambda$ by writing \emph{external terms}, $e$. The abstract syntax of external terms is shown in Figure \ref{syntax-EL} (we will give various concrete desugarings in Sec. \ref{external-terms}). %We will describe useful syntactic desugarings atop this syntax as we go on and review recent techniques that permit modularly introducing  desugarings like these in Sec. \ref{desugaring}. Our main focus here is on semantic extensions. 
The static and dynamic semantics are specified simultaneously as a \emph{bidirectionally typed translation semantics}, i.e. the key judgements take the form:\[\esynX{e}{\st\moutput}{\iota\moutput} ~~~~~\text{and}~~~~~ \eanaX{e}{\st}{\iota\moutput}\]
\noindent
These are pronounced ``$e$ (synthesizes / analyzes against) type $\sigma$ and has  translation $\iota$ under typing context $\Upsilon$ and tycon context $\Phi$''. Note that our specifications in this paper are intended to be algorithmic: we indicate ``outputs'' when introducing judgement forms by \emph{mode annotations}, $\moutput$; these are not part of the judgement's syntax. In particular, note that the type is an ``output'' only for the synthetic judgement.

This basic separation of the EL and IL is commonly used for full-scale language specifications, e.g. the Harper-Stone semantics for Standard ML \cite{Harper00atype-theoretic}. The internal language is purposely kept small, e.g. defining only binary products, to simplify metatheoretic reasoning and compilation. The EL then specifies various useful higher-level constructs, e.g. record types, by translation to the IL. In @$\lambda$, the EL builds in only function types. All other external constructs are defined in the \emph{tycon context}, described starting in the Sec. \ref{types}.

We choose bidirectional typechecking, also sometimes called \emph{local type inference} \cite{Pierce:2000:LTI:345099.345100}, for two main reasons. The first is once again to justify the practicality of our approach: local type inference is increasingly being used in modern languages (e.g. Scala \cite{OdeZenZen01}) because it eliminates the need for type annotations in many situations while remaining decidable in more situations than whole-function type inference and providing what are widely perceived to be higher quality error messages \cite{journals/jfp/JonesVWS07}. Secondly, it  gives us a clean way to reuse the generalized introductory form, $\eintro{\st}{\es}$, and its associated desugarings, at many types \cite{TSLs}. For example, regular string types can use standard string literal syntax.

Unlike the Harper-Stone semantics, where external and internal terms were governed by a common type system, in @$\lambda$ each external type, $\st$, maps onto an internal type, $\tau$, called the \emph{type translation} of $\st$. This mapping is specified by the  type translation judgement, $\vdash_\Phi \st \leadsto \tau$, which will be described in Sec. \ref{sec:type-translations}. 
For this reason, this specification style may also be compared to specifications for the first stage of a type-directed compiler, e.g. the TIL compiler for Standard ML \cite{tarditi+:til-OLD}, here lifted ``one level up'' into the semantics of the language itself. As we will see, type safety follows from a property analagous to a correctness condition that arises in typed compilers. Modular reasoning will be based on holding the type translation of $\st$ abstract ``outside'' the tycon.

External typing contexts $\Upsilon$ map variables to types, so we also need the judgement $\vdash_\Phi \Upsilon \leadsto \Gamma$ to construct a $\Gamma$ that maps the same variables to corresponding type translations.

@@

\paragraph{Static Language}
\begin{figure}[t]
\small
\hspace{-5px}\input{syntax-table-SL}
\caption{Syntax of the static language (SL). Metavariable $\sx$ ranges over static term variables, $\kalpha$ and $\kvar$ over kind variables and $n$ over natural numbers.}\vspace{-5px}
\label{syntax-SL}
\end{figure}
The workhorse of @$\lambda$ is the \emph{static language}, which itself forms a typed lambda calculus where 
%External terms are classified by (external) \emph{types}. 
\emph{kinds}, $\kappa$, classify \emph{static terms}, $\sigma$.  The syntax of the SL is given in Figure \ref{syntax-SL}. The portion of the SL covered by the first row of kinds and first row of static terms, some of which are elided for concision, forms an entirely standard  functional programming language consisting of total functions, universal quantification over kinds, inductive kinds, and products and sums. The reader can consider these as forming a total subset of ML or a simply-typed subset of Coq and we will assume standard conveniences from such languages (e.g. let bindings, pattern matching and inference of type parameters when applying polymorphic functions) are available in examples for concision. The semantics we assume also directly follows \cite{pfpl}, so we also omit the details of this core here.  Only three new  kinds are needed for the SL to serve its role as the language used to   control typing and translation of the EL: $\kty$ (Sec. \ref{types}), $\kity$ (Sec. \ref{sec:type-translations}) and $\kitm$ (Sec. \ref{sec:introop}).

The kinding judgement takes the form $\sofkX{\st}{\kappa\moutput}$, where $\kDelta$ and $\kGamma$ are analagous to $\iDelta$ and $\iGamma$ and analagous kind and kinding context formation judgements $\kDelta \vdash \kappa$ and $\kDelta \vdash \kGamma$ are  defined. All such contexts in @$\lambda$ are identified up to exchange and contraction and obey weakening \cite{pfpl}. The natural number $n$ is used as a technical device in our semantics to ensure that the forms shown as being indexed by $n$ in the syntax only arise in a controlled manner internally to prevent ``out of bounds'' issues, as we will discuss; they would have no corresponding concrete syntax so $n$ can be assumed $0$ in user-defined terms.

The dynamic semantics of static terms is defined as a structural operational semantics by a stepping judgement $\sstep{\st}{\argEnv}{\st\moutput}$, a value judgement $\sval{\st}{\argEnv}$ and an error judgement $\serr{\st}{\argEnv}$. Here, $\argEnv$ ranges over \emph{argument environments}, which we will  return to when considering opcons in Sec. \ref{external-terms}. The multi-step judgement $\smanystep{\st}{\argEnv}{\st\moutput}$ is the reflexive, transitive closure of the stepping judgement. The normalization judgement $\seval{\st}{\argEnv}{\st'}$ is defined iff $\smanystep{\st}{\argEnv}{\st'}$ and $\sval{\st'}{\argEnv}$. %${\st}{\memD}{\memG}{\aCtx}{\st\moutput}{\memD\moutput}{\memG\moutput}$, where $\memD$, $\memG$ and $\aCtx$ are also technical devices that will be described later; they too can be ignored from the perspective of user code. We write $\st \Downarrow \st'$ iff $\snorm{\st}{\emptyset}{\cdot}{{\rightharpoonup}; \emptyset; \emptyset}{\st'}{\emptyset}{\cdot}$. \emph{Static values} are normalized static terms. Normalization can also raise an error (to indicate a type error in an external term, or a problem in a tycon definition, as we will discuss), indicated by the judgement $\serrX{\st}$. We omit error propagation rules.%We will refer to the relevant rules as we proceed. 
\begin{figure}[t]
\small
$\begin{array}{lrcl}
\textbf{tycons} & c & ::= & \rightharpoonup ~|~ \tc ~|~ \keyw{other}[m]\\
\textbf{tycon contexts} & \Phi & ::= & \cdot ~|~ \Phi, \tcdef{\tc}{\psi}{\theta}\\
\textbf{tycon structures} & \theta & ::= & \tcstruct{\st}{\omega} \\
\textbf{opcon structures} & \omega & ::= & \tcstructn{\st} ~|~ \tcstructc{\omega}{op}{\st}\\
\textbf{tycon sigs} & \psi & ::= & \tcsig{\kappa}{\chi}\\
\textbf{opcon sigs}& \chi & ::= & \introsig{\kappa} ~|~ \opsigS{\chi}{op}{\kappa}\\
\end{array}$
\caption{Syntax of tycons. Metavariables $\tc$ and $\opname{op}$ range over user-defined tycon and opcon names, respectively, and $m$ ranges over natural numbers.}
\label{syntax-TC}\vspace{-8px}
\end{figure}
\begin{figure*}\begin{mathpar}
\small
\inferrule[k-parr]{
    \sofkX{\st}{\kprod{\kty}{\kty}}
}{
    \sofkX{\sty{\rightharpoonup}{\st}}{\kty}
}

\inferrule[k-ty]{
    \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\theta} \in \Phi\\
    \sofkX{\st}{\ktyidx}
}{
    \sofkX{\sty{\tc}{\st}}{\kty}
}

\inferrule[k-otherty]{
    \sofkX{\sttyidx}{\keyw{Nat} \times \kity}
}{
    \sofkX{\sty{\keyw{other}[m]}{\sttyidx}}{\kty}
}\vspace{-10px}
\end{mathpar}
\caption{Kinding rules for types, which take the form $\sty{c}{\sttyidx}$ where $c$ is a tycon and $\sttyidx$ is the type index.}
\label{fig:types}\vspace{-10px}
\end{figure*}

@@

% (k-ty)

>>>>

In this paper, we begin from first principles by specifying a language, @$\lambda$, in a style that follows many full-scale languages: as a bidirectionally typed translation semantics. Only the $\rightarrow$ type constructor (tycon) is built in; all other external tycons (we show constrained strings and a variation on record types) are defined by extending a \emph{tycon context}. Each tycon defines the semantics of its associated term-level \emph{opcons} (e.g. row projection) using a static language where types and translations are values. The semantics guarantees \emph{type safety} and, unlike other approaches, \emph{conservativity}: that all \emph{tycon-specific invariants} that can be established in a ``closed world'' will be conserved in the ``open world''. Type system providers need not provide mechanized proofs. Instead, these follow from a form of translation validation that uses type abstraction to check that tycons maintain \emph{translation independence}, analagous to the representation independence maintained by ML-style module systems.
\end{abstract}

@@

Universality properties (e.g. ``Turing completeness'') are often enough to guarantee that an embedding that preserves a desirable fragment's dynamic semantics can be constructed, but an isomorphic embedding must also preserve the static semantics and, if defined, performance bounds specified using a cost semantics. This is not always possible. For example, in $\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\keyw{1}}\,{\times}\,{+}\}$, it is impossible to introduce record types as a library because you need the row projection operators, in ML written $\opname{\#}\conclbl{lbl}$, one for each of the infinite set of row labels $\conclbl{lbl}$ ($\opname{\#}$ itself is an \emph{operator constructor}). %You can only use $\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\keyw{1}}\,{\times}\,{+}\}$ as a \emph{target of compilation} for a language with record types (they can be implemented using nested binary products,  lists, or other types). %(and attempting to simulate records via a hash table over strings plus a refinement type system to check that the lookup operation will not fail will not satisfy a cost semantics prescribing $\mathcal{O}(1)$ lookup; see Sec. \ref{prior-work}).

%Embeddings are also sometimes too \emph{complex}, as measured by the cost of the extralinguistic computations that are needed to map in and out of the embedding and, if these  must be performed mentally by programmers, considering various human factors. %We will discuss specific examples below.
%When an embedding is too complex, abstraction providers have a few options, discussed in Sec. \ref{prior-work}. When an embedding is not possible, however, or when these options are  insufficient, 
%providers are often compelled to introduce these fragments by extending an existing language, thereby forming a new \emph{dialect}. To save effort, they may do so by forking existing artifacts or leverage tools like compiler generators or language frameworks, also reviewed in Sec. \ref{prior-work}. %Indeed, the proliferation of language dialects constructed for this reason might be taken as an evidence that the core language is not as ``general'', when considered comprehensively, as might be hoped. 
Each time a fragment must be introduced directly, rather than as a library, a new \emph{dialect} of the language arises. Within the ML lineage, dialects that go beyond ``core ML'' abound:%, showing that this is not an isolated example:
%Reynolds, in a remark that recalls the ``Turing tarpit'' of Perlis \cite{Perl82a}, summarizes the issue \cite{Reynolds94anintroduction}: 
%\begin{quote}
%To say that any reasonable function can be expressed by some program is not to say that it can be expressed by the most reasonable program. It is clear that the language requires a novel programming style. Moreover, it is likely that certain important functions cannot be expressed by their most efficient algorithms.
%\end{quote}
%

\begin{enumerate}
\vspace{-7px}
%\compresslist
\item 
\textbf{General Purpose Fragments:} 
Many variations on product types, for example, exist: 
$n$-ary tuples, 
labeled tuples, 
records (identified up to reordering), 
records with width and depth subtyping \cite{Cardelli:1984:SMI:1096.1098}, 
records with update operators\footnote{The Haskell wiki notes that ''No, extensible records are not implemented in GHC. The problem is that the record design space is large, and seems to lack local optima. [...] As a result, nothing much happens.'' \cite{GHCFAQ}} \cite{ocaml-manual}, 
records with mutable fields \cite{ocaml-manual}, 
%field delegation \cite{atlang-gpce14} \todo{gpce submission} 
and 
records with ``methods'' (i.e. pure objects \cite{TSLs}). 
 Sum-like types are also exposed variously: 
standard datatypes, 
open datatypes \cite{conf/ppdp/LohH06,journals/toplas/MillsteinBC04}, 
polymorphic variants \cite{ocaml-manual} and 
exception types \cite{harper1997programming}. Combinations of these manifest themselves as class-based object systems \cite{ocaml-manual}.

\item
\textbf{Specialized Fragments:} Fragments that track specialized static invariants are also frequently introduced in dialects, e.g. for distributed programming \cite{Murphy:2007:TDP:1793574.1793585}, reactive programming \cite{mandel2005reactiveml}, authenticated data structures \cite{Miller:2014:ADS:2535838.2535851}, databases \cite{Ohori:2011:MSM:2034773.2034815},  units of measure \cite{conf/cefp/Kennedy09} and regular string sanitation \cite{sanitation-psp14}, amongst many other examples. %\verb|sprintf| in the OCaml dialect statically distinguishes format strings from strings.% All of these are implemented as dialects of existing languages, presumably because a strong encoding was not feasible.
\vspace{-3px}
\item
\textbf{Foreign Fragments:} A safe and natural foreign function interface (FFI) can be valuable (particularly given this proliferation of dialects). This requires enforcing the type system of the foreign language in the calling language. %Using  a FFI that does not do this can lead to safety issues, even when both languages are separately known to be safe. 
%Safe FFIs generally require direct extensions to the language. 
For example, MLj builds in a safe FFI to Java \cite{Benton:1999:IWW:317636.317791}.
\end{enumerate}
\vspace{-5px}

This \emph{dialect-oriented} state of affairs is unsatisfying. %Language designers are burdened with needing to understand the complete metatheory of every fragment they add to their language to make sure it is not incompatible with existing fragments, so languages change rarely. This decreases the impact of many potentially useful innovations designed by fragment providers (e.g. many academic researchers) by making them unavailable to programmers. 
While programmers can choose from dialects supporting, e.g., a principled approach to distributed programming, or one that builds in support for statically reasoning about units of measure, one that supports both fragments may not be available. Using different dialects separately for different components of a program is untenable: components written in different dialects cannot always interface safely (i.e. a safe FFI, item 3 above, is needed between every pair of dialects). %And as just mentioned, composing dialects is non-trivial.

These problems do not arise for fragment expressed as an isomorphic embedding (i.e. as a library) because modern \emph{module systems} can enforce abstraction barriers that ensure that the isomorphism needs only to be established in the ``closed world'' of the module. For example, a module defining sets in ML can hold the representation of sets abstract, ensuring that any invariants necessary to maintain the isomorphism need only be maintained by the functions in the module (e.g. uniqueness, if using a list representation). These then continue to hold no matter which other modules are in use by a client \cite{harper1997programming}. Other languages provide similar forms of \emph{abstract data types} to localize reasoning \cite{liskov1974programming}. %Mechanisms that can help decrease the complexity of an embedding without violating abstraction barriers are thus valuable, and we will lead into our work in Sec. \ref{prior-work} by summarizing them.

Mechanisms are needed that make it possible to define and  reason in a similarly modular, localized manner about  direct extensions to the semantics of a language. For example, if a language is extended with \emph{regular string types} as described in \cite{sanitation-psp14}, all terms having a regular string type like $\sty{\tcvar{rstr}}{\concrx{[A-Z]+ \digit\digit\digit\digit}}$ should continue to behave as strings in the regular language indicated no matter which other extensions are in use by a client. %Such a mechanism could ultimately be integrated directly into the language, blurring the distinction between fragments and libraries and decreasing the need for new dialects.% Importing fragments that introduce new semantics   would be as easy and reasonable as importing a new module is in ML today. %In the limit, the community could rely on modularly mechanized metatheory and compiler correctness results.% rather than requiring heroic efforts from individual research groups that consider an entire language at once. %Recent work has shown progress on modularly introducing new concrete syntax, reviewed in Sec. \ref{desugaring}. Our focus is on the problem of introducing new semantics.

\paragraph{Contributions} In this paper, we take foundational steps toward this goal by  constructing a simple but surprisingly powerful core calculus, @$\lambda$ (the ``actively typed'' lambda calculus). %Despite its minimality, it can host a variety of practical semantic extensions like those described above, while maintaining strong metatheoretic guarantees and, crucially, providing modular reasoning properties. %a calculus introduced briefly in recent work on \emph{active type constructors}  \cite{atlang-gpce14}\todo{TR? Arxiv?}, reviewed in Sec. \ref{overview}. 
Its semantics are structured like those of many modern languages, consisting of an \emph{external language} (EL) governed by a {typed translation semantics} targeting a much simpler \emph{internal language} (IL). 
Rather than building in a monolithic set of external type constructors, however, the  semantics are indexed by a \emph{tycon context}. Each tycon, e.g. $\tcvar{lprod}$ defining labeled products or $\tcvar{rstr}$ defining regular strings, determines the semantics of its associated opcons (e.g. $\opname{\#}$ for row projection, or $\opname{conc}$ for concatenation) via \emph{static functions}, i.e. functions written in a \emph{static language} (SL), where types and translations are values.

All translation invariants maintained by these opcons in any ``closed world'', e.g. the regular string invariant just mentioned, are guaranteed to be maintained in any further extended tycon context, i.e. in the ``open world'', due to a simple static check that maintains \emph{translation independence} between tycons using type abstraction in the IL, the same fundamental mechanism underlying representation independence in ML-style module systems. As in ML, mechanized specifications and proofs are not needed.%, which provides a form of type-level computation. %The authors demonstrate the expressive power of this technique primarily with an implementation and a number of examples of fragments as libraries, and their core calculus .% (unlike systems that treat the type system as a ``bag of rules'', where non-determinism can arise).

% We will begin by giving an overview of  @$\lambda$ in Sec. \ref{atlam}, introducing our two main examples, one defining labeled product types (i.e. record types maintaining an ordering for simplicity) with a functional update operator, and the other regular string types, based on a recent core calculus style specification \cite{sanitation-psp14}. In Sec. \ref{types}, we detail how tycons are defined and how types arise from tycons. We then describe how tycons control the semantics of their term-level operator constructors (opcons) in Sec. \ref{external-terms}.  We next give the key metatheoretic properties of the calculus in Sec. \ref{metatheory}, including \emph{type safety} and a key modularity result, which we call  \emph{conservativity}: any invariants that can be established about all values of a type under \emph{some} tycon context (i.e. in some  ``closed world'') are conserved in any further extended tycon context (i.e. in the ``open world'').  Interestingly, type system providers need not provide mechanized proofs to maintain such guarantees. Instead, the approach we take relies on type abstraction in the internal language. As a result, we are able to use the same parametricity results that underly modular reasoning in simply-typed languages like ML to reason modularly about typed language fragments. We conclude with  related and future work in Sec. \ref{prior-work}.

\section{Overview of @$\lambda$}\label{atlam}\label{overview}
\begin{figure}[t]
\small
\hspace{-5px}\input{syntax-table-IL}
\caption{Syntax of {$\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\iunit}\,{\times}\,{+}\}$}, our internal language (IL). Metavariable $x$ ranges over term variables and $\alpha$ and $t$ both range over type variables.}
\label{syntax-IL}
\vspace{-10px}
\end{figure}
\paragraph{External Language}
Programmers interface with @$\lambda$ by writing \emph{external terms}, $e$. The abstract syntax of external terms is shown in Figure \ref{syntax-EL} and we will introduce various concrete desugarings in Sec. \ref{external-terms}. %We will describe useful syntactic desugarings atop this syntax as we go on and review recent techniques that permit modularly introducing  desugarings like these in Sec. \ref{desugaring}. Our main focus here is on semantic extensions. 
The semantics are specified as a \emph{bidirectionally typed translation semantics}, i.e. the key judgements have the form:\[\esynX{e}{\st\moutput}{\iota\moutput} ~~~~~\text{and}~~~~~ \eanaX{e}{\st}{\iota\moutput}\]
\noindent
These are pronounced ``$e$ (synthesizes / analyzes against) type $\sigma$ and has  translation $\iota$ under typing context $\Upsilon$ and tycon context $\Phi$''. Note that our specifications in this paper are intended to be algorithmic: we indicate ``outputs'' when introducing judgement forms by \emph{mode annotations}, $\moutput$. Note that the type is an ``output'' only for the synthetic judgement.

This separation of the EL from the IL is commonly used for full-scale language specifications, e.g. the Harper-Stone semantics for Standard ML \cite{Harper00atype-theoretic} or the semantics of Wyvern \cite{TSLs}. The internal language is purposely kept small, e.g. defining only binary products, to simplify metatheoretic reasoning and compilation. The EL then specifies various useful higher-level constructs, e.g. record types, by translation to the IL. In @$\lambda$, the EL builds in only function types. All other external constructs are defined in the \emph{tycon context}, $\Phi$, described starting in the Sec. \ref{types}.

We choose bidirectional typechecking, also sometimes called \emph{local type inference} \cite{Pierce:2000:LTI:345099.345100}, for two main reasons. The first is once again to justify the practicality of our approach: local type inference is increasingly being used in modern languages (e.g. Scala \cite{OdeZenZen01}) because it eliminates the need for type annotations in many situations while being simpler and decidable in more situations than whole-function type inference and providing what are widely perceived to be higher quality error messages \cite{journals/jfp/JonesVWS07}. Secondly, it  gives us a clean way to reuse the generalized introductory form, $\eintro{\st}{\es}$, and its associated desugarings, at many types \cite{TSLs}. For example, regular string types can use standard string literal syntax.

Unlike the Harper-Stone semantics, where external and internal terms were governed by a common type system, in @$\lambda$ each external type, $\st$, maps onto an internal type, $\tau$, called the \emph{type translation} of $\st$. This mapping is specified by the  type translation judgement, $\vdash_\Phi \st \leadsto \tau$, which will be described in Sec. \ref{sec:type-translations}. For example, regular string types will translate to internal type abbreviated $\keyw{str}$, and labeled product types will translate to nested binary product types, though we will emphasize that there are other valid choices.

This specification style may thus also be compared to specifications for the first stage of a type-directed compiler, e.g. the TIL compiler for Standard ML \cite{tarditi+:til-OLD}, here lifted ``one level up'' into the semantics of the language itself. As we will see, type safety follows from a property analagous to a correctness condition that arises in typed compilers. Modular reasoning will be based on holding the type translation of $\st$ abstract ``outside'' the tycon.

External typing contexts $\Upsilon$ map variables to types, so we also need the judgement $\vdash_\Phi \Upsilon \leadsto \Gamma$.

@@

\paragraph{Static Language}
\begin{figure}[t]
\small
\hspace{-5px}\input{syntax-table-SL}
\caption{Syntax of the static language (SL). Metavariable $\sx$ ranges over static term variables, $\kalpha$ and $\kvar$ over kind variables and $n$ over natural numbers.}\vspace{-5px}
\label{syntax-SL}
\end{figure}
The main novelty of @$\lambda$ is the \emph{static language}, which itself forms a typed lambda calculus where we use the word 
%External terms are classified by (external) \emph{types}. 
\emph{kind}, $\kappa$, to refer to the ``types'' of \emph{static terms}, $\sigma$.  The syntax of the SL is given in Figure \ref{syntax-SL}. The portion of the SL covered by the first row of kinds and static terms, some of which are elided for concision, forms an entirely standard total functional programming language consisting of total functions, universal quantification over kinds, inductive kinds, and products and sums  \cite{pfpl}. The reader can consider these as forming a total subset of ML \cite{harper1997programming} or a simply-typed subset of Coq \todo{citation}and we will assume standard conveniences from such languages (e.g. let bindings, pattern matching and inference of type parameters when applying polymorphic functions) in examples for concision. Only three new  kinds are needed for the SL to serve its role as the language used to   control typing and translation of the EL: $\kty$, classifying types (Sec. \ref{types}), $\kity$, classifying \emph{translational internal types} (Sec. \ref{sec:type-translations}) and $\kitm$, classifying \emph{translational internal terms} (Sec. \ref{sec:introop}).

The kinding judgement takes the form $\sofkX{\st}{\kappa\moutput}$, where $\kDelta$ and $\kGamma$ are analagous to $\iDelta$ and $\iGamma$ and analagous kind and kinding context formation judgements $\kDelta \vdash \kappa$ and $\kDelta \vdash \kGamma$ are  defined. All such contexts in @$\lambda$ are identified up to exchange and contraction and obey weakening \cite{pfpl}. The natural number $n$ is used as a technical device in our semantics to ensure that the forms shown as being indexed by $n$ arise in a controlled manner to prevent ``out of bounds'' issues, as we will discuss; they would have no corresponding concrete syntax so $n$ can be assumed $0$ in user-defined terms.

The dynamic semantics of static terms is defined as a structural operational semantics by a stepping judgement $\sstep{\st}{\argEnv}{\st\moutput}$, a value judgement $\sval{\st}{\argEnv}$ and an error raised judgement $\serr{\st}{\argEnv}$. $\argEnv$ ranges over \emph{argument environments}, which we will  return to when considering opcons in Sec. \ref{external-terms}. The multi-step judgement $\smanystep{\st}{\argEnv}{\st\moutput}$ is the reflexive, transitive closure of the stepping judgement. The normalization judgement $\seval{\st}{\argEnv}{\st'}$ is defined iff $\smanystep{\st}{\argEnv}{\st'}$ and $\sval{\st'}{\argEnv}$. %${\st}{\memD}{\memG}{\aCtx}{\st\moutput}{\memD\moutput}{\memG\moutput}$, where $\memD$, $\memG$ and $\aCtx$ are also technical devices that will be described later; they too can be ignored from the perspective of user code. We write $\st \Downarrow \st'$ iff $\snorm{\st}{\emptyset}{\cdot}{{\rightharpoonup}; \emptyset; \emptyset}{\st'}{\emptyset}{\cdot}$. \emph{Static values} are normalized static terms. Normalization can also raise an error (to indicate a type error in an external term, or a problem in a tycon definition, as we will discuss), indicated by the judgement $\serrX{\st}$. We omit error propagation rules.%We will refer to the relevant rules as we proceed.

@@

% (k-tc)
